### 2018-07-10 source/Ch4-Qweights/paper.tex
backup

\chapter{Analysing the Variance of Gain Solutions}\label{chapter.paper}
\minitoc
\input{source/Ch4-Qweights/paperdefs.tex}
\

@@

\pg
Interferometers sample Fourier modes of the sky brightness distribution corrupted by instrumental and atmospheric effects rather than measuring the sky brightness directly.
This introduces two problems for astronomers to invert: calibration and imaging. Both of these problems are ill-conditioned.

@@

\pg
The problem of calibration is what concerns us in this section. It consists of estimating and correcting for instrumental errors (which includes effects such as antenna pointing errors, but also the phase-delays caused by ionospheric activity, troposphere, etc). 
 {Calibration consists of solving for gain estimates, where a gain models the relationship between the electromagnetic field of an astrophysical source and the voltage that an antenna measures for this source. Because measurements are noisy, calibration often involves some fine-tuning of solution intervals, to ensure that the solutions are well-constrained while the solution intervals stay as small as signal-to-noise allows.} The calibration inverse problem involves three competing statistical effects: thermal noise in the measurements, true gain variability, and {sky} model incompleteness. If gain solutions {are sought individually for each measurement}, then calibration estimates will be dominated by thermal noise, and will not adequately describe the actual gains. Similarly, if a single gain estimate is fitted to too many measurements, the intrinsic gain variability will be ``averaged out''; for example, a choice of time and frequency interval that is too large will cause the solver to estimate a constant gain while the underlying function varies quickly, thereby missing much of the gain structure. This will introduce error which will be correlated in time and frequency. This occurs, for example, when solving for ionospheric phase delays: in the most extreme case, where the solution interval is significantly larger than the scale of ionospheric fluctuations, its varying phase can average out to zero over the interval in time and frequency. Finally, if the model being fitted is incomplete, unmodeled physical flux will likely be absorbed unpredictably into both the gain solutions and the residual visibilities: this absorption of physical flux into gain solutions is known as source suppression \citep[see][and references therein]{2014MNRAS.439.4030G,2013MNRAS.435..597K}.

@@

In practice, it is reasonable to assume that gain variation is generally slower than some given scale: we can then reduce the noise of our gain estimates by finding a single gain solution for a small number of measurements, assuming that the underlying gain variation is very small and stable over short intervals. This is \emph{generally} a valid hypothesis, but the specific value for the variation scale can be contentious. Indeed, while the noise level can often be treated as constant throughout an observation, the gain variability itself is generally not constant: there will be time periods where the gains will tend to remain constant for longer, and others where variability will be very quick. This means that, for any choice of calibration interval, some gain estimates will be better than others, and almost all could have been improved (at a cost to others) by a different choice of time (and frequency) intervals.

@@

\pg
In this section, we derive our first fundamental result: the Cov-Cov relationship, Eq. \ref{eq.covcov.matrix}, which describes how the statistics of residual visibilities (and thus the antenna calibration solutions, henceforth ``gains") relate to the statistics of the image plane, i.e. of images made using the associated visibilities. The dimensions of the matrices (denoted by boldface capital letters) and vectors (denoted by boldface lowercase letters) used in this paper are given in Table \ref{tab.variables}, along with the scalar numbers used to denote specific dimensions. All other variables are scalars.

@@

Note that the only direction-dependent terms in the above are $s_d$ and $\kappa_b^d$, which are both inside $\phi_{bb'}$ (for both $b=b'$ and $b \ne b'$). {By making the approximation that the Fourier kernels of different sources are orthogonal (i.e. that $\kappa_b^d \kappa_{b'}^{d'} = (\kappa_b^d)^2 \delta_{bb'}$)}\footnote{{This hypothesis is equivalent to assuming that the sky is dominated by distant point sources, where ``distant" means that the sources are multiple PSF Full-Width Half-Maximum apart}} we can write:

@@

{where we have now limited our formalism to the case where the sky is dominated by distant point-like sources.}

@@

This is our fundamental result: {assuming unpolarised emission coming from distant point sources and normally-distributed thermal noise}, it gives a direct relationship between the covariance of the residual visibilities and the covariance of the residual image-pixel values. We thus call it the Cov-Cov relationship. It describes the statistical properties of the image-plane as the result of a convolution process {changing an average noise level at different points in the image-plane, allowing} us to describe the behaviour of variance and covariance in the image. Let us focus on the first

@@

\pg
This $\deltu\deltv$ space warrants further discussion: Fig. \ref{plot.psf.uv.dudv} shows, for a given $uv$-track (Fig. \ref{plot.psf.uv.dudv}b), both the corresponding $\deltu\deltv$ domain (Fig. \ref{plot.psf.uv.dudv}c) and point-spread function (Fig. \ref{plot.psf.uv.dudv}a). {The symmetric, negative $uv$-track is treated as a separate track, and thus ignored. This means that we do not fully constrain the noise-PSF (since the covariance matrix of the symmetric track is simply the Hermitian of the first), but we do not seek to constrain it in this section, but rather to show that our results hold.} We can see that the $\delta u \delta v$-tracks are symmetrical about the origin. The $\delta u \delta v$ space corresponding to a given $uv$-track can thus be most concisely described as a ``filled $uv$-track", with its boundaries defined by the ends of the $uv$-track. The set of $\muelF_{bb'}$, each of which maps one value of the covariance matrix to a fringe in the image-plane, would then characterise a PSF equivalent for the noise distribution, which we refer to as the noise-PSF. In our formalism, the only sources of statistical effects in the field are calibration errors and thermal noise. The average variance in all pixels will be given by the diagonal of the covariance matrix and the thermal noise, provided that it truly follows a normal distribution. The only effects which will cause the variance in the image-plane to vary from one pixel to the next are those mapped onto the covariance fringes, i.e. such position-dependent variance fluctuations will be caused by correlated gain errors, which are spurious signal introduced by erroneous gain estimates. Assuming all sources in the field are point-like and distant, then these variance fluctuations will follow a specific distribution, convolved to every source in the field.

@@

\pg
{This dirty image simply consists of performing a direct Fourier transform (i.e. without using a Fast Fourier Transform algorithm) on simulated visibilities corresponding to these three point sources. We now perform a similar test as above on this field. Firstly, we ``apply" gain errors to these visibilities by multiplying our model with our residual gain errors. This allows us to find 2000 realisations of residual visibilities, and find the variance for each pixel across these realisations. This gives us the simulated noise-map, shown in Fig. \ref{imag.simu-3sources.noisemap}a. Secondly, we perform a DFT from the differential Fourier plane to the $(l,m)$ plane as before, assigning one cell of $\Cov{\matGains}$ to each point of the differential Fourier plane. This time, however, $\phi_{bb'}^d$ is not simply unity for all points in the differential Fourier plane. Instead, it is calculated for the three-point-source model, and applied for each point. This gives us the predicted noise-map in Fig. \ref{imag.simu-3sources.noisemap}b. Finally, Fig. \ref{imag.simu-3sources.noisemap}c shows the absolute value of the difference between the two images. We see that there is some structure present in these residuals: this is expected, as the PSF of the sources in the dirty imags clearly overlap. We are thus not quite in the regime where emission is fully spatially incoherent. Nevertheless, our predictions hold to better than $5\%$ accuracy. }

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{images/Ctime6400-noisePSFandDirty-uncorr.png}
\caption{\label{imag.simu-3sources.noisemap} {Noise-map of sky with correlated gain errors and three point sources. The colourbars of (a), (b) and (c) have dimensionless units, while that of (d) is in Jansky. Note the presence of structure in the residuals (c): these show the limits of our hypothesis that sources are spatially incoherent.}}
\end{figure}

@@

\pg
$\Rmat{\cellindex}$ corresponds to a matrix containing all the residual visibilities within one calibration cell $\cellindex$, i.e. for $\tnuIndeix \in \cellindex$ where $\gvect_{\cellindex} = \const$ Let us define $n_\cellindex$ as the number of elements in each calibration cell. The residual variance sample matrix can now be built by multiplying the residual visibility matrix with its Hermitian conjugate:
\begin{align}
\matVV             =& \bm{R}^H \bm{R} \label{eq.matvv.definition} %\\
\end{align}
Note that we do this because it allows us to turn a single noise realisation $\epsilon$ into a statistical quantity $\sigma$.
We can relate $\matVV$ to the variance of individual antenna gains:
\begin{align}
\matVV
            =& \sum_{\tnuIndeix\in\cellindex} \left(\sum_{d,d'} s_d \Kmat{d,\tnuIndeix} \left( \VarmatC \right)^H \Kmat{d,\tnuIndeix}^H s_{d'} \Kmat{d',\tnuIndeix} \left( \VarmatC \right) \Kmat{d',\tnuIndeix}^H + \I\sigma^2 \right)
\end{align}
To reach this point, in Eq. \ref{whocares}, we made the hypothesis that the sky brightness distribution is dominated by spatially incoherent emission. Applying this hypothesis again here, we can make the approximation that the cross-terms in the sum over $d,d'$ average to zero: $\sum_{d,d'\ne d}\approx 0$. We then have:
\begin{align}
\matVV\approx& \sum_\tnuIndeix \left(\sum_d s_d^2 \Kmat{d,\tnuIndeix} \left( \VarmatC \right)^H  \left( \VarmatC \right) \Kmat{d,\tnuIndeix}^H + \I\sigma^2\right)\\
            =& \left(\VarmatC\right)^2\hadam \left(\underbrace{\sum_{\tnuIndeix}\sum_d s_d^2 \kvect{d,\tnuIndeix}\kvect{d,\tnuIndeix}^H}_{\bydefn \mathcal{S}}\right) + n_\cellindex\I\sigma^2\\
\VarmatC    =& \sqrt{\mathcal{S}^{\hadam-1}\left(\matVV-n_\cellindex\I\sigma^2\right)}
\end{align}
where $\hadam$ denotes the Hadamard or entrywise product and $\kvect{}=\diag{\Kmat{}}$. % If $\matVV$ is diagonal, the gains are uncorrelated and the above can be written as:
Thus, $\matVV$ allows us to estimate the variance of each antenna and for each calibration cell by using all the visibilities pointing to that antenna within that calibration cell. With this information, we can then rebuild the baseline-dependent matrix, having improved our sampling by a factor of $n_{\mathrm{ant}}$.

@@

\pg
In this section, we show the effect of adaptive quality-based weighting on real data. The dataset used in this section is a single sub-band from an 8-hour LOFAR observation centred on the Extended Groth Strip ($\alpha$=14:19:17.84,$\delta$=52:49:26.49). The observation was performed on August 28th, 2014. The subband includes 8 channels of width 24.414 kHz each, for a total bandwidth ranging from 150.2 to 150.4 MHz. The data have been averaged in time to 1 data point per second. The data was calibrated using Wirtinger calibration \citep[see][and references therein]{2014arXiv1410.8706T,2015MNRAS.449.2668S} and a sky model consisting only of a nearby calibrator source, 3C295. A reference image (a cutout of which is shown in Fig. \ref{image.3c295.goodcal}) was made by calibrating the data according to best practice for LOFAR survey data: 1 calibration solution per 8 seconds and per 4 channels. {The residual data was then corrected by the gain solutions and imaged} using Briggs weighting (robust=0), pixel size of $1.5''$, and deconvolved using the default devoncolution algorithm in DDFacet (Tasse et al., in press).

>>>>

\chapter{Analysing the Variance of Gain Solutions}\footnote{This section is based on \citetads{2017arXiv171100421B}, which was published as part of this PhD.}\label{chapter.paper}
\minitoc
\input{source/Ch4-Qweights/paperdefs.tex}
\

@@

\pg
Interferometers sample Fourier modes of the sky brightness distribution corrupted by instrumental and atmospheric effects rather than measuring the sky brightness directly.
This introduces two problems\footnote{Which are simply an artificial split of a single original problem; see \cref{sec.CalibImagery} for more details.} for astronomers to invert: calibration and imaging. Both of these problems are ill-conditioned.

@@

\pg
The problem of calibration is what concerns us in this section. It consists of estimating and correcting for instrumental errors (which includes effects such as antenna pointing errors, but also the phase-delays caused by ionospheric activity, troposphere, etc). 
{Calibration consists of solving for gain estimates, where a gain models the relationship between the electromagnetic field of an astrophysical source and the voltage that an antenna measures for this source. Because measurements are noisy, calibration often involves some fine-tuning of solution intervals, to ensure that the solutions are well-constrained while the solution intervals stay as small as signal-to-noise allows.} The calibration inverse problem involves three competing statistical effects: thermal noise in the measurements, true gain variability, and {sky} model incompleteness. If gain solutions {are sought individually for each measurement}, then (unless the observed field is \textit{exceptionally} bright) calibration estimates will be dominated by thermal noise, and will not adequately describe the actual gains. Similarly, if a single gain estimate is fitted to too many measurements, the intrinsic gain variability will be ``averaged out''; for example, a choice of time and frequency interval that is too large will cause the solver to estimate a constant gain while the underlying function varies quickly, thereby missing much of the gain structure. This will introduce error which will be correlated in time and frequency. This occurs, for example, when solving for ionospheric phase delays: in the most extreme case, where the solution interval is significantly larger than the scale of ionospheric fluctuations, its varying phase can average out to zero over the interval in time and frequency. Finally, if the model being fitted is incomplete, unmodeled physical flux will likely be absorbed unpredictably into both the gain solutions and the residual visibilities: this absorption of physical flux into gain solutions is known as source suppression \citep[see][and references therein]{2014MNRAS.439.4030G,2013MNRAS.435..597K}.

@@

In practice, it is reasonable to assume that there exist a scale at which gain variation is acceptably low: we can then reduce the noise of our gain estimates by finding a single gain solution for a small number of measurements, assuming that the underlying gain variation is very small and stable over short intervals. This is \emph{generally} a valid hypothesis, but the specific value for the variation scale can be contentious. Indeed, while the noise level can often be treated as constant throughout an observation, the gain variability itself is generally not constant: there will be time periods where the gains will tend to remain constant for longer, and others where variability will be very quick. This means that, for any choice of calibration interval, some gain estimates will be better than others, and almost all could have been improved (at a cost to others) by a different choice of time (and frequency) intervals.

@@

\pg
In this section, we derive our first fundamental result: the Cov-Cov relationship, Eq. \ref{eq.covcov.matrix}, which describes how the statistics of residual visibilities (and thus the antenna calibration solutions, henceforth ``gains") relate to the statistics of the image plane, i.e. of images made using the associated visibilities. The dimensions of the matrices (denoted by boldface capital letters) and vectors (denoted by boldface lowercase letters) used in this work are given in Table \ref{tab.variables}, along with the scalar numbers used to denote specific dimensions. All other variables are scalars.

@@

Note that the only direction-dependent terms in the above are $s_d$ and $\kappa_b^d$, which are both inside $\phi_{bb'}$ (for both $b=b'$ and $b \ne b'$). {By making the approximation that the Fourier kernels of different sources are orthogonal (i.e. that $\kappa_b^d \kappa_{b'}^{d'} = (\kappa_b^d)^2 \delta_{bb'}$)}\footnote{{This hypothesis is equivalent to assuming that the sky is dominated by well-separated point sources, where we take ``well-separated" to mean that the sources are multiple PSF Full-Width Half-Maximum apart}} we can write:

@@

{where we have now limited our formalism to the case where the sky is dominated by sufficiently-separated point-like sources.}

@@

This is our fundamental result: {assuming unpolarised emission coming from well-separated point sources and normally-distributed thermal noise}, it gives a direct relationship between the covariance of the residual visibilities and the covariance of the residual image-pixel values. We thus call it the Cov-Cov relationship. It describes the statistical properties of the image-plane as the result of a convolution process {changing an average noise level at different points in the image-plane, allowing} us to describe the behaviour of variance and covariance in the image. Let us focus on the first

@@

\pg
This $\deltu\deltv$ space warrants further discussion: Fig. \ref{plot.psf.uv.dudv} shows, for a given $uv$-track (Fig. \ref{plot.psf.uv.dudv}b), both the corresponding $\deltu\deltv$ domain (Fig. \ref{plot.psf.uv.dudv}c) and point-spread function (Fig. \ref{plot.psf.uv.dudv}a). {The symmetric, negative $uv$-track is treated as a separate track, and thus ignored. This means that we do not fully constrain the noise-PSF (since the covariance matrix of the symmetric track is simply the Hermitian of the first), but we do not seek to constrain it in this section, but rather to show that our results hold.} We can see that the $\delta u \delta v$-tracks are symmetrical about the origin. The $\delta u \delta v$ space corresponding to a given $uv$-track can thus be most concisely described as a ``filled $uv$-track", with its boundaries defined by the ends of the $uv$-track. The set of $\muelF_{bb'}$, each of which maps one value of the covariance matrix to a fringe in the image-plane, would then characterise a PSF equivalent for the noise distribution, which we refer to as the noise-PSF. In our formalism, the only sources of statistical effects in the field are calibration errors and thermal noise. The average variance in all pixels will be given by the diagonal of the covariance matrix and the thermal noise, provided that it truly follows a normal distribution. The only effects which will cause the variance in the image-plane to vary from one pixel to the next are those mapped onto the covariance fringes, i.e. such position-dependent variance fluctuations will be caused by correlated gain errors, which are spurious signal introduced by erroneous gain estimates. Assuming all sources in the field are point-like and distant, then these variance fluctuations will follow a specific distribution, convolved with every source in the field.

\pg
By the same token, while an image made with visibilities containing only Gaussian noise will have a flat variance, the noise-PSF will correlates the noise between pixels - and the noise-correlation will be given exactly by the PSF. Noise in interferometric images is therefore always correlated between pixels -- removing this correlation would require a full image deconvolution. In a sense, this is expected: this is what the PSF, in a very fundamental way, encodes.

@@

\pg
{This dirty image simply consists of performing a direct Fourier transform (i.e. without using a Fast Fourier Transform algorithm) on simulated visibilities corresponding to these three point sources. We now perform a similar test as above on this field. Firstly, we ``apply" gain errors to these visibilities by multiplying our model with our residual gain errors. This allows us to find 2000 realisations of residual visibilities, and find the variance for each pixel across these realisations. This gives us the simulated noise-map, shown in Fig. \ref{imag.simu-3sources.noisemap}a. Secondly, we perform a DFT from the differential Fourier plane to the $(l,m)$ plane as before, assigning one cell of $\Cov{\matGains}$ to each point of the differential Fourier plane. This time, however, $\phi_{bb'}^d$ is not simply unity for all points in the differential Fourier plane. Instead, it is calculated for the three-point-source model, and applied for each point. This gives us the predicted noise-map in Fig. \ref{imag.simu-3sources.noisemap}b. Finally, Fig. \ref{imag.simu-3sources.noisemap}c shows the absolute value of the difference between the two images. We see that there is some structure present in these residuals: this is expected, as the PSF of the sources in the dirty imags clearly overlap. We are thus not quite in the regime where emission is from well-separated point sources. Nevertheless, our predictions hold to better than $5\%$ accuracy. }

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{images/Ctime6400-noisePSFandDirty-uncorr.png}
\caption{\label{imag.simu-3sources.noisemap} {Noise-map of sky with correlated gain errors and three point sources. The colourbars of (a), (b) and (c) have dimensionless units, while that of (d) is in Jansky. Note the presence of structure in the residuals (c): these show the limits of our hypothesis that sources are well-separated.}}
\end{figure}

@@

\pg
$\Rmat{\cellindex}$ corresponds to a matrix containing all the residual visibilities within one calibration cell $\cellindex$, i.e. for $\tnuIndeix \in \cellindex$ where $\gvect_{\cellindex} = \const$ Let us define $n_\cellindex$ as the number of elements in each calibration cell. The residual variance sample matrix can now be built by multiplying the residual visibility matrix with its Hermitian conjugate:
\begin{align}
\matVV             =& \bm{R}^H \bm{R} \label{eq.matvv.definition} %\\
\end{align}
Note that we do this because it allows us to turn a single noise realisation $\epsilon$ into a statistical quantity $\sigma$.
We can relate $\matVV$ to the variance of individual antenna gains:
\begin{align}
\matVV
            =& \sum_{\tnuIndeix\in\cellindex} \left(\sum_{d,d'} s_d \Kmat{d,\tnuIndeix} \left( \VarmatC \right)^H \Kmat{d,\tnuIndeix}^H s_{d'} \Kmat{d',\tnuIndeix} \left( \VarmatC \right) \Kmat{d',\tnuIndeix}^H + \I\sigma^2 \right)
\end{align}
To reach this point, in Eq. \ref{whocares}, we made the hypothesis that the sky brightness distribution is dominated by emission from well-separated point sources. Applying this hypothesis again here, we can make the approximation that the cross-terms in the sum over $d,d'$ average to zero: $\sum_{d,d'\ne d}\approx 0$. We then have:
\begin{align}
\matVV\approx& \sum_\tnuIndeix \left(\sum_d s_d^2 \Kmat{d,\tnuIndeix} \left( \VarmatC \right)^H  \left( \VarmatC \right) \Kmat{d,\tnuIndeix}^H + \I\sigma^2\right)\\
            =& \left(\VarmatC\right)^2\hadam \left(\underbrace{\sum_{\tnuIndeix}\sum_d s_d^2 \kvect{d,\tnuIndeix}\kvect{d,\tnuIndeix}^H}_{\bydefn \mathcal{S}}\right) + n_\cellindex\I\sigma^2\\
\VarmatC    =& \sqrt{\mathcal{S}^{\hadam-1}\left(\matVV-n_\cellindex\I\sigma^2\right)}
\end{align}
where $\hadam$ denotes the Hadamard or entrywise product and $\kvect{}=\diag{\Kmat{}}$. % If $\matVV$ is diagonal, the gains are uncorrelated and the above can be written as:
Thus, $\matVV$ allows us to estimate the variance of each antenna and for each calibration cell by using all the visibilities pointing to that antenna within that calibration cell. With this information, we can then rebuild the baseline-dependent matrix, having improved our sampling by a factor of $n_{\mathrm{ant}}$.

@@

\pg
In this section, we show the effect of adaptive quality-based weighting on real data. The dataset used in this section is a single sub-band from an 8-hour LOFAR observation centred on the Extended Groth Strip ($\alpha$=14:19:17.84,$\delta$=52:49:26.49). The observation was performed on August 28th, 2014. The subband includes 8 channels of width 24.414 kHz each, for a total bandwidth ranging from 150.2 to 150.4 MHz. The data have been averaged in time to 1 data point per second. The data was calibrated using Wirtinger calibration \citep[see][and references therein]{2014arXiv1410.8706T,2015MNRAS.449.2668S} and a sky model consisting only of a nearby calibrator source, 3C295. A reference image (a cutout of which is shown in Fig. \ref{image.3c295.goodcal}) was made by calibrating the data according to best practice for LOFAR survey data: 1 calibration solution per 8 seconds and per 4 channels. {The residual data was then corrected by the gain solutions and imaged} using Briggs weighting (robust=0), pixel size of $1.5''$, and deconvolved using the default devoncolution algorithm in DDFacet \citepads{2017arXiv171202078T}.
